# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

First, we set the working directory where the extracted csv-file was placed. This file is loaded into the object "raw"

```{r load the data}
setwd("D:/Documents/R/DataScience/git/RepData_PeerAssessment1")
raw <- read.csv("activity.csv")
```

Second, the column containing the dates is formated as such for later recognition of weekdays

```{r process/transform the data}
tidy <- raw
tidy$date <- as.Date(raw$date, format = "%Y-%m-%d")
```


## What is mean total number of steps taken per day?

Here is a histogram of the total number of steps taken each day:

```{r histogram steps per day}
library(plyr)
steps.day <- ddply(tidy, .(date), summarize, sum=sum(steps))
hist(steps.day$sum, xlab = "number of steps taken each day", ylab = "number of days within the 2-month-range", main = "Histogram (ignoring missing values)")
```

Here is the mean and median total number of steps taken per day:

```{r mean and median}
steps.day.mean <- ddply(tidy, .(date), summarize, mean=mean(steps), median=median(steps))
steps.day.mean
```


## What is the average daily activity pattern?

The daily activity pattern averages across all days; for the time series plot, missing values have to be treated, here simply ignored.

```{r time series plot of 5min intervals over all days}
steps.interval <- ddply(tidy, .(interval), summarize, average=mean(steps, na.rm=TRUE))
plot(steps.interval, type = 'l', xlab = "Interval", ylab = "Average number of steps")
max.int <- steps.interval$interval[steps.interval$average==max(steps.interval$average)]
```

The 5-minute interval `r max.int` contains the maximum number of steps.


## Imputing missing values

```{r number of NA}
missing <- sum(is.na(tidy$steps))
```

There are `r missing` missing values, i.e. number of rows with NAs, in the dataset.

A new dataset is created, named tidy2, with missing date replaced by the mean for that 5-min interval.

```{r replacing NA}
tidy2 <- tidy
for (i in 1:dim(tidy)[1]) {
  if (is.na(tidy2[i,1])==TRUE) {
    tidy2[i,1] <- steps.interval$average[steps.interval$interval==tidy2[i,3]]
  }
}
```

Repeat the previous step done on tidy with tidy2: 

Here is a histogram of the total number of steps taken each day (missing values replaced by mean for that 5-min interval)

```{r histogram steps per day2}
library(plyr)
steps.day2 <- ddply(tidy2, .(date), summarize, sum=sum(steps))
hist(steps.day2$sum, xlab = "number of steps taken each day", ylab = "number of days within the 2-month-range", main = "Histogram (missing values replaced by 5-min average)")
```

Here is the mean and median total number of steps taken per day (missing values replaced by mean for that 5-min interval)

```{r mean and median2}
steps.day.mean2 <- ddply(tidy2, .(date), summarize, mean=mean(steps), median=median(steps))
steps.day.mean2
```

The NA's affected only eight days, but those entirely (8 time 288 equals the total of 2304 missing values). Therefore, the mean over a day was not useful to replace the NA's. Consequently, only those 8 days were affected, and all of them received identical values, i.e. for each of the 288 5-min intervals the average of the 53 valid days.  
In the histogram, the middle bar grew by 8.  
In the mean/median table, these 8 days received identical mean/median replacing the previous NAs; the other 53 days remaining unchanged.


## Are there differences in activity patterns between weekdays and weekends?

A new factor variable is created in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday day or weekend day.

```{r weekdays}
tidy2$wd.type <- weekdays(tidy2$date)
tidy2$wd.type <- gsub("Montag", "weekday", tidy2$wd.type)
tidy2$wd.type <- gsub("Dienstag", "weekday", tidy2$wd.type)
tidy2$wd.type <- gsub("Mittwoch", "weekday", tidy2$wd.type)
tidy2$wd.type <- gsub("Donnerstag", "weekday", tidy2$wd.type)
tidy2$wd.type <- gsub("Freitag", "weekday", tidy2$wd.type)
tidy2$wd.type <- gsub("Samstag", "weekend", tidy2$wd.type)
tidy2$wd.type <- gsub("Sonntag", "weekend", tidy2$wd.type)
tidy2$wd.type <- as.factor(tidy2$wd.type)
```

The daily activity pattern averages across all days, separating between weekday days and weekend days. 

```{r time series plot of 5min intervals over all weekdays and weekends respectively}
steps.interval.wd <- ddply(tidy2, .(interval, wd.type), summarize, average=mean(steps))
library(lattice)
xyplot(average ~ interval | wd.type, data = steps.interval.wd, layout = c(1,2), type = 'l', xlab = "Interval", ylab = "Average number of steps")

```

